Rapid and Accurate Global PTM Discovery (G-PTM-D) Using Post-Acquisition Spectral Calibration and Defined Mass Windows

### 2017-02-28 paper.tex
fixes

Before describing the proposed changes to G-PTM-D, we introduce the database search mode definitions we use throughout. A standard \textit{narrow-mass} search considers matching theoretical peptides to spectra if the peptide mass and the identified precursor mass are equal, within some tolerance (i.e. 5ppm, or 2.1 Da). An \textit{wide-mass} search matches peptide fragment masses to fragmentation spectra while allowing large differences (i.e. $\pm 500$ Da) between the observed precursor mass and the peptide mass. An even less restrictive \textit{open-mass} search, matches peptides from a database to  fragmentation spectra without taking the precursor mass into account. We propose alternatives to these search modes, namely a \textit{notch search} and an \textit{interval search} that fall under a broad category of \textit{mass-window} searches. The notch search is an extension of the narrow-mass search: it allows exact matches (within a tolerance) to specific mass differences between the theoretical peptide and observed precursor masses. The interval search is an extension of the wide-mass search: specific problematic mass differences are excluded from allowed matches. The smaller the fraction of the mass space, and the narrower the mass window per search element, the faster the search will run (as fewer searches are executed per mass spectrum). Figure~\ref{fgr:differentSearchModes} provides a convenient way to visualize the various search modes.

We propose an extension to G-PTM-D (Enhanced G-PTM-D) that addresses the long search times of the wide-mass search, and the limitations of the standard final narrow-mass search. The wide-mass search is replaced by a notch search in the discovery phase, using notches centered at the monoisotopic mass differences introduced by known modifications. This change improves the specificity and significantly decreases search. The final narrow-mass search is also replaced by a notch search, further increasing the number of confidently identified peptides.

We recommend using the mass-window search in other procedures as well, beyond the G-PTM-D workflow. Its efficacy is demonstrated in context of a search for novel modifications, which is usually based on a wide-mass or an open-mass search. We demonstrate that using a combination of specialized notch and interval searches produces superior identification and discovery results.

@@

\subsection{Enhanced G-PTM-D}

@@

Even though adducts arising from experimental artefacts are not ultimately interesting to biologists, not including them as an option decreases the total number of identified proteins and PTMS. A curated set of notches and the corresponding modifications is given in Appendix A. Using an off-the shelf database of modification is not advisable. The database augmentation step is like the procedure described in the original G-PTM-D work: modifications with appropriate mass are added to the protein database, so that subsequent searches may include them as variable localized modifications. 
Since the XML database is annotated with PTMs from the UniProt PTM list, it is a natural first idea to use it in the augmentation step as well. It has a few significant drawbacks: Some mass values do not correspond to actual mass shifts, lability of modifications is not taken into account, and many PTMs are missing, including adducts and glycosylations.
The final search with the augmented database is conducted with notches: the final notch search is specific to each dataset: we only recommend including the common missed monoisotopic mass values in the final search: see the following section for justification. Newly identified localized PTMs are now readily identified.

By employing the novel notch search strategy instead of the wide-mass search in the G-PTM-D process, using only the modifications listed in Uniprot, the number of confidently identified PTMs increased by 8\%. The overall search time dropped significantly, from 35 hours to 4 hours for all datasets, see Table~\ref{my-label}. Using an expanded modification list that includes chemical modifications such as adducts and large mass PTMs such as glycosylations, the overall identification rate increased from 164,697 to 179,345 peptide-spectral matches (PSMs), which in turn increased the number of modified peptides by an additional 20\%. We identified hundreds of glycosylated peptides in these unenriched samples, with many of these modifications exceeding 1000 Daltons. Algorithmic improvements in the G-PTM-D procedure itself, such as searching for combinations of modifications and for modifications with neutral losses make it an excellent tool for PTM discovery.

>>>>

Before describing the proposed changes to G-PTM-D, we introduce the database search mode definitions we use throughout. A standard \textit{narrow-mass} search considers matching theoretical peptides to spectra if the peptide mass and the identified precursor mass are equal, within some tolerance (i.e. 5ppm, or 2.1 Da). A \textit{wide-mass} search matches peptide fragment masses to fragmentation spectra while allowing large differences (i.e. $\pm 500$ Da) between the observed precursor mass and the peptide mass. An even less restrictive \textit{open-mass} search, matches peptides from a database to  fragmentation spectra without taking the precursor mass into account. We propose alternatives to these search modes, namely a \textit{notch search} and an \textit{interval search} that fall under a broad category of \textit{mass-window} searches. The notch search is an extension of the narrow-mass search: it allows exact matches (within a tolerance) to specific mass differences between the theoretical peptide and observed precursor masses. The interval search is an extension of the wide-mass search: specific problematic mass differences are excluded from allowed matches. The smaller the fraction of the mass space, and the narrower the mass window per search element, the faster the search will run (as fewer searches are executed per mass spectrum). Figure~\ref{fgr:differentSearchModes} provides a convenient way to visualize the various search modes.

We propose an extension to G-PTM-D (Enhanced G-PTM-D) that addresses the long search times of the wide-mass search, and the limitations of the standard final narrow-mass search. The wide-mass search is replaced by a notch search in the discovery phase, using notches centered at the monoisotopic mass differences introduced by known modifications. This change improves the specificity and significantly decreases search time. The final narrow-mass search is also replaced by a notch search, further increasing the number of confidently identified peptides.

We recommend using the mass-window search in other procedures as well, beyond the G-PTM-D workflow. Its efficacy is demonstrated in context of a search for novel modifications, which is usually based on a wide-mass or an open-mass search. We demonstrate that a combination of specialized notch and interval searches produces superior identification and discovery results.

@@

\subsection{Enhanced G~-PTM~-D}

@@

Even though adducts arising from experimental artefacts are not ultimately interesting to biologists, not including them as an option decreases the total number of identified proteins and PTMS. A curated set of notches and the corresponding modifications is given in Appendix A. Using an off-the shelf database of modification is not advisable. The database augmentation step is like the procedure described in the original G~-PTM~-D work: modifications with appropriate mass are added to the protein database, so that subsequent searches may include them as variable localized modifications. 
Since the XML database is annotated with PTMs from the UniProt PTM list, it is a natural first idea to use it in the augmentation step as well. It has a few significant drawbacks: Some mass values do not correspond to actual mass shifts, lability of modifications is not taken into account, and many PTMs are missing, including adducts and glycosylations.
The final search with the augmented database is conducted with notches: the final notch search is specific to each dataset: we only recommend including the common missed monoisotopic mass values in the final search: see the following section for justification. Newly identified localized PTMs are now readily identified.

By employing the novel notch search strategy instead of the wide-mass search in the G~-PTM~-D process, using only the modifications listed in Uniprot, the number of confidently identified PTMs increased by 8\%. The overall search time dropped significantly, from 35 hours to 4 hours for all datasets, see Table~\ref{my-label}. Using an expanded modification list that includes chemical modifications such as adducts and large mass PTMs such as glycosylations, the overall identification rate increased from 164,697 to 179,345 peptide-spectral matches (PSMs), which in turn increased the number of modified peptides by an additional 20\%. We identified hundreds of glycosylated peptides in these unenriched samples, with many of these modifications exceeding 1000 Daltons. Algorithmic improvements in the G~-PTM~-D procedure itself, such as searching for combinations of modifications and for modifications with neutral losses make it an excellent tool for PTM discovery.
